Supporting Dynamic Data Partitioning in Parallel Programs

### 2012-11-01 doc/types-pldi2013/intro.tex
more intro

Recently machine architecture, particularly at the high
performance end of the spectrum, has undergone a revolution.  The
latest supercomputers are now composed of heterogeneous processors
and deep memory hierarchies.  Current programming systems for these
machines have elaborate features for describing parallelism, but 
few abstractions for describing the organization of data.  However, as the architecture
of distributed memory has become increasingly complex, the placement and movement of dat has
become a first order programming concern.

@@

Previous work has explored two approaches to expressing the organization of data
in distributed memory parallel programs.
One class of work uses fully static analysis 
with no runtime overhead but limited expressivity.  Two recent examples, 
Sequoia \cite{Fatahalian06} and 
Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a 
mechanism to statically partition the heap into a tree of collections 
of data.  The two designs are different in many aspects, but agree that there is
a single tree-shaped partitioning of data that must be fully checked statically 
(see Section~\ref{sec:related}).  The
second approach relies on fully dynamic analysis of program data that doesn't
limit expressivity but incurs runtime overhead;  examples 
include transactional memory \cite{Harris05} and thread-level speculation \cite{Steffan00}.

Our own experience writing high-performance 
applications in both Sequoia and the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that a
single, static data partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions must be dynamically computed.
Furthermore, applications often need multiple, simultaneous partitions of
the same data, providing different views on to that data.  Alternatively, 
the fully dynamic approach permits arbitrary partitioning, but requires 
centralized control logic to perform conflict detection between every pair 
of parallel computations, resulting in runtime overhead that isn
not scalable on distributed-memory machines.

In this work, we present a type system and operational semantics for
Legion\cite{Legion12}.  Legion is a programming model that supports multiple,
dynamic partitions of program data which can result in aliasing.  We prove
the soundness of our type system and show that through a combination of
static and dynamic analysis we can implement it in a way that minimizes dynamic
runtime checks and results in low actual runtime overhead.  We then show 
that the soundness of our type system permits a hierarchical,
distributed scheduling algorithm that enables Legion programs to scale to 
high performance on current supercomputers.

Legion presents {\em logical regions} as an
abstraction for capturing {\em locality} (data that will be used together,
and therefore should be colocated) and {\em independence} (disjoint data
that may be operated on in parallel).  Logical regions have the following
properties that allow programmers to dynamically communicate information
about the structure of program data to Legion:
\begin{itemize}
\item  logical regions are first-class values in Legion
and may be dynamically allocated and stored in data structures;

@@

\item  a logical region may be dynamically partitioned in multiple different ways.
\end{itemize}
Through multiple partitions
of a logical region, a given datum may belong to multiple different regions.  
For example, if a region $R$ is disjointly partitioned three distinct ways 
and each partition assigns every element in $R$ to a
subregion (partitions need not be total in Legion), then every element
in $R$ is included in three different subregions.  This is why we use
the term {\em logical} regions: language-level regions serve to name
sets of data but do not imply anything about physical layout or placement
in the memory hierarchy.\footnote{There
is a separate system of {\em physical regions} at runtime that hold
copies of the data in the language-level logical regions.  Details can
be found in \cite{Legion12}.}

While data in Legion is organized into a hierarchy of logical region forests, 
computation is organized into a hierarchy of tasks.  Logical regions and tasks interact
through a system of privileges specifying operations that a task
can perform on a logical region argument (read, write, or reduce).  
Tasks are only permitted to access logical regions for which they 
possess privileges.  Privileges are given to a task by their calling task
or by creating new logical regions.  The Legion programming model 
mandates a functional restriction: a sub-task can only be passed a subset
of the privileges possesed by its parent task.

To enforce this restriction, our type system relies on a combination of static 
and dynamic analyses.  The key observation is that dynamic
analysis is very easy and relatively cheap when done at the granularity of
regions instead of individual heap locations.  To that end, our type system supports
static verification of all pointer dereferences and passing of region privileges at 
task calls boundaries.  The Legion runtime then leverages its knowledge of logical
region aliasing to perform dynamic safety checks for the unpacking
and packing of logical regions to/from the heap with regards to task privileges.  
These checks are then amortized over all the uses of the logical region 
which significantly reduces their cost.

To enable parallelism, Legion again relies on a combination of static and dynamic
analysis to prove that two tasks operate on independent sets of data.  If two tasks
cannot be proven independent statically, the Legion runtime performs dynamic dependence
tests to detect region aliasing.  Since these tests are performed at the granularity of
regions instead of individual heap locations, they are significantly less costly.
In addition to supporting parallel execution for independent tasks, Legion also enables
programmer specified parallel execution in the presence of aliasing.  For example, two tasks 
accessing aliased data may only need to be serialized
with respect to each other (e.g. run atomically) or are permitted to run
simultaneously with the programmer using separate synchronization mechanisms
to control access (e.g. locks).  To support these relaxed access paradigms 
Legion provides {\em cohrence modes} on logical regions.  For each region
argument to a task, a programmer specifies the coherence in which other tasks
can access aliased data while the task is running.  We present an operational semantics
for coherence modes and show that it is sound.

Using our proofs of the soundess of the privilegs and coherence systems, we then 
prove a third result: if two {\em siblings} (tasks that
have the same immediate parent task) $t_1$ and $t_2$ are 
{\em non-interfering} (have no ordering requirements for correctness), 
then any descendant of $t_1$ in the task hierarchy is non-interfering
with any descendant of $t_2$.  This theorem is the basis for correct, distributed
task scheduling in a Legion implementation, which is crucial for
performance; a centralized scheduler would be a bottleneck
because of the large communication latencies in the target
class of machines.

@@

The rest of the paper is organized as follows.
We begin in Section~\ref{sec:example} with an example program that illustrates
a typical style of programming in Legion and motivates the need for multiple, dynamically computed partitions of
a region that introduce aliased data.  We define Core Legion, a small language suitable 
for our formal development, in Section~\ref{sec:legioncore}.
The next four sections each state and prove one of our main results and contributions:
\begin{itemize}
\item We prove the soundness of Legion's privileges system (Section~\ref{sec:soundness}).

\item We show that if expressions $e_1$ and $e_2$ are {\em non-interfering} (can be executed in parallel), then subexpressions
$e_1'$ of $e_1$ and $e_2'$ of $e_2$ are also non-interfering (Section~\ref{sec:scheduling}).  
%This result is the basis
%for a hierarchical, distributed scheduler in the Legion implementation which is crucial for high performance
%on the target class of machines.
%; on the target class of machines, any centralized
%scheduler would be a serious bottleneck.

\item We give experimental evidence for the Legion design.  On three real-world applications, we show that 
dynamic region pointer checks would be expensive, justifying checking this aspect of the type system statically.
We also show that the cost of checking for non-interference of region privileges is low, showing that a much more expressive
and dynamic language with aliasing is not incompatible with high performance (Section~\ref{sec:evaluation}).

\end{itemize}
Finally, Section~\ref{sec:related} discusses the most closely related work and
Section~\ref{sect:conclusion} concludes.

>>>>

Machine architecture, particularly at the high
performance end of the spectrum, has recently undergone a revolution.  The
latest supercomputers are now composed of heterogeneous processors
and deep memory hierarchies.  Current programming systems for these
machines have elaborate features for describing parallelism, but 
few abstractions for describing the organization of data.  However, as the architecture
of distributed memory has become increasingly complex, the placement and movement of data has
become a first order programming concern.

@@

Previous work has explored two approaches to expressing the organization of data
in distributed memory parallel programs.
One class of work uses fully static analysis 
with no runtime overhead but limited expressivity.  Two recent examples, 
Sequoia \cite{Fatahalian06} and 
Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a 
mechanism to statically partition the heap into a tree of collections 
of data.  The two designs are different in many aspects, but agree that there is
a single tree-shaped partitioning of data that must be fully checked statically 
(see Section~\ref{sec:related}).  The
second approach relies on fully dynamic analysis of program data that doesn't
limit expressivity but incurs runtime overhead and centralized control logic that is
difficult to scale to distributed memory machines;  examples 
include transactional memory \cite{Harris05} and thread-level speculation \cite{Steffan00}.

Our own experience writing high-performance 
using the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that a
single, static data partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions must be dynamically computed.
Furthermore, applications often need multiple, simultaneous partitions of
the same data, providing different views on to that data.

In this work, we present the design of Legion \cite{Legion12},
a parallel programming model that efficiently supports multiple, dynamic partitions
of data. Legion's {\em logical regions} are an
abstraction capturing {\em locality} (data that will be used together,
and therefore should be colocated) and {\em independence} (disjoint data
that may be operated on in parallel).  Logical regions have 
properties that allow programmers to dynamically communicate information
about the structure of program data to Legion:
\begin{itemize}
\item  logical regions are first-class values in Legion
and may be dynamically allocated and stored in data structures;

@@

\item  a logical region may be dynamically partitioned in multiple different ways;

\item programmers give logical regions {\em privilege} and {\em coherence} properties that
describe how the region's data may be used by computations.
\end{itemize}

While data in Legion is organized in a hierarchy of logical region forests, 
computation is organized in a hierarchy of tasks.  Logical regions and tasks interact
through a system of privileges specifying operations that a task
can perform on a logical region argument (read, write, or reduce).  
Tasks can only access logical regions for which they 
possess privileges.  Tasks acquire privileges from their calling task
or by creating new logical regions.  A key restriction is that 
a sub-task can only be passed a subset of the privileges of its parent task.

Through multiple partitions of a logical region, a given datum may
belong to multiple different regions.  Thus, allowing multiple
simultaneous partitions of the same data introduces
aliasing---distinct regions may have overlapping data.\footnote{This
  is why we use the term {\em logical} regions: language-level regions
  name sets of data but do not imply a physical layout or placement in
  the memory hierarchy. A separate system of {\em physical regions} at
  runtime holds copies of data in the language-level logical regions
  \cite{Legion12}.}  We believe permitting very flexible aliasing is
necessary, but it is also presents a major challenge. For example, to
detect that tasks can execute in parallel it is sufficient to prove
that they work on disjoint data.

Legion also provides {\em coherence} modes for regions, which allows
programmers to permit parallel execution even in cases where tasks in
fact do access the same data in conflicting ways.  The programmer may
specify that tasks conflicting on a region  must execute serially (in program
order, the default), {\em atomically} with respect to each other, or even
{\em simultaneously} if the programmer uses their own synchronization mechanisms
to control access (e.g., locks).

To efficiently enforce restrictions that guarantee the safety and
correctness of program executions for uses of first-class regions,
their privilege and coherence properties, and arbitrary region
aliasing, the Legion runtime relies on a combination of static and
dynamic checks.  A key theme that runs through the design is that very
hard static analysis problems (e.g., alias analysis) are often both easy and
inexpensive if they can be done dynamically at the granularity of regions instead of individual
heap locations, while checks that must be done on individual heap locations are only
cheap if done statically.  Explicating and evaluating the design of Legion's
static and dynamic semantics and its implications is the main topic of this paper,
which we present as a series of contributions:
\begin{itemize}
\item We present a type system for {\em Core Legion} programs that statically
  verifies the safety of individual pointer references and region
  privileges at task call boundaries (Section~\ref{subsec:coretypes}).
  A subtle issue arises in checking privileges when regions are
  stored in and later retrieved from heap data structures.  The type
  system provides enough static information that the Legion runtime
  can leverage its dynamic information about region aliasing to
  perform dynamic privelege checks guaranteeing safety.

\item We present a parallel operational semantics for Core Legion that
  may be of independent interest.  This semantics is compositional,
  hierarchical, and asynchronous, reflecting the way such programs
  actually execute on the hardware.  However, the semantics also
  directly models conflicting memory operations between tasks in cases of relaxed
  coherence, making explicit the boundary between deterministic and
  possible non-deterministic behavior (Section~\ref{subsec:opsemantics}).

\item We prove the soundness of Legion's privelege system (Section~\ref{sec:soundness}).

\item We show that if expressions $e_1$ and $e_2$ are {\em non-interfering} (can be executed in parallel), then subexpressions
$e_1'$ of $e_1$ and $e_2'$ of $e_2$ are also non-interfering (Section~\ref{sec:scheduling}).  
This result is the basis
for a hierarchical, distributed scheduler in the Legion implementation which is crucial for high performance
on the target class of machines;  any form of centralized scheduling would be a bottleneck because of the
large communication latencies in the target class of machines.
%; on the target class of machines, any centralized
%scheduler would be a serious bottleneck.

\item We give experimental evidence for the Legion design.  On three
  real-world applications, we show that dynamic region pointer checks
  would be expensive, justifying checking this aspect of the type
  system statically.  We also show that the cost of region aliasing
  checks is low, showing that a much more expressive and dynamic
  language with aliasing is not incompatible with both high
  performance and safety (Section~\ref{sec:evaluation}).

\end{itemize}

@@

%The rest of the paper is organized as follows.
%We begin in Section~\ref{sec:example} with an example program that illustrates
%a typical style of programming in Legion and motivates the need for multiple, dynamically computed partitions of
%a region that introduce aliased data.  We define Core Legion, a small language suitable 
%for our formal development, in Section~\ref{sec:legioncore}.
%Finally, Section~\ref{sec:related} discusses the most closely related work and
%Section~\ref{sect:conclusion} concludes.

### 2012-11-01 doc/types-pldi2013/legion-types.tex
more intro

\begin{abstract}
Many distributed-memeory parallel applications dynamically partition 
data, but current language designs either 
require fully static partitioning or rely on entirely dynamic techniques.
We describe Legion, a programming system with a combination
of language features and static and dynamic checks to support 
dynamic data partitions with minimal overhead.
We describe the novel aspects of the Legion design, prove the soundness of the Legion type system,
 and show Legion type checking improves performance by up to 71\% by eliding provably
safe memory checks.  Furthermore, we show how the language design and type safety
enables a hierarchical, distributed runtime scheduling algorithm, and how
a system of data coherence annotations improves expressivity without compromising
soundness. We report results for three real-world applications running
on distributed memory machines, achieving up to 62.5X speedup on 96 GPUs 
on the Keeneland supercomputer.
%Legion is a programming model that introduces
%{\em logical regions} as a mechanism for targeting large heterogeneous machines 
%with complex memory hierarchies.  Logical regions are first-class 
%values in Legion, and programmers may {\em partition} regions in multiple different
%ways into subregions.  We present
%a type system with privileges and coherence properties for reasoning about logical regions.  
%We prove the soundness of the type system and use it to show
%that all memory accesses in Legion are safe.  Furthermore, we prove that
%type soundness enables a scalable, hierarchical
%scheduling algorithm for Legion programs.  We show that an implementation of the type checker
%for Legion programs improves performance up to 71\% by eliding
%memory checks the type system proves safe.  We also demonstrate that an implementation of the Legion 
%runtime based on the hierarchical scheduling algorithm scales well for
%three real-world applications.

>>>>

\begin{abstract}
Many distributed-memeory parallel applications dynamically partition 
data, but current language designs either 
require fully static partitioning or rely on expensive dynamic checks.
We describe Legion, a programming system with a combination
of language features and static and dynamic checks to support 
dynamic data partitions with minimal overhead.
We describe the novel aspects of the Legion design, prove the soundness of the Legion type system,
 and show Legion type checking improves performance by up to 71\% by eliding provably
safe memory checks.  Furthermore, we show how the language design and type safety
enables a hierarchical, distributed runtime scheduling algorithm, and how
a system of data coherence annotations improves expressivity without compromising
soundness. We report results for three real-world applications running
on distributed memory machines, achieving up to 62.5X speedup on 96 GPUs 
on the Keeneland supercomputer.
%Legion is a programming model that introduces
%{\em logical regions} as a mechanism for targeting large heterogeneous machines 
%with complex memory hierarchies.  Logical regions are first-class 
%values in Legion, and programmers may {\em partition} regions in multiple different
%ways into subregions.  We present
%a type system with privileges and coherence properties for reasoning about logical regions.  
%We prove the soundness of the type system and use it to show
%that all memory accesses in Legion are safe.  Furthermore, we prove that
%type soundness enables a scalable, hierarchical
%scheduling algorithm for Legion programs.  We show that an implementation of the type checker
%for Legion programs improves performance up to 71\% by eliding
%memory checks the type system proves safe.  We also demonstrate that an implementation of the Legion 
%runtime based on the hierarchical scheduling algorithm scales well for
%three real-world applications.
